### 2025-08-30 chapters/background.tex
chapter 3 complete
review complete

RISC-V specifies three hierarchical privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{Lee2019}. Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the core Trusted Computing Base (TCB)—within M-mode. The SM is responsible for managing enclaves, enforcing strict memory isolation, and mediating access to sensitive operations. By anchoring the TCB at this highest privilege level, Keystone minimizes the trusted surface exposed to potential attackers.

A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10. PMP empowers M-mode software to define fine-grained access permissions for physical memory regions, specifying which lower privilege levels (U and S) may read, write, or execute code therein~\cite{keystone2025how}. This hardware-enforced mechanism forms the cornerstone of Keystone’s memory isolation strategy. Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{Lee2019}.

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.48\linewidth}
  \centering
  \begin{tikzpicture}[font=\small]
    \draw[fill=green!10] (0,2) rectangle (3,5) node[midway]{\textbf{Secure World}};
    \draw[fill=red!10] (5,2) rectangle (8,5) node[midway]{\textbf{Normal World}};
    \draw[fill=gray!20] (2,0) rectangle (6,1.5) node[midway]{\textbf{Secure Monitor}};
    \draw[->,thick] (3,3.5) -- (5,3.5) node[midway,above]{\textbf{SMC}};
  \end{tikzpicture}
  \caption{TrustZone's dual-world architecture with SMC-managed transitions.}
  \label{fig:trustzone}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\linewidth}
  \centering
  \begin{tikzpicture}[font=\small]
    \node[above] at (3,4) {\textbf{Enclave Page Cache (EPC)}};
    \draw[thick] (0,0) rectangle (6,4);

@@

\begin{table}[htbp]
\centering
\caption{Comparison of three TEE implementations.}
\label{tab:tee-comparison}
\small
%\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{3cm}p{2.5cm}p{2.5cm}p{2.5cm}@{}}
\toprule
\textbf{Feature} & \textbf{Intel SGX} & \textbf{ARM TrustZone} & \textbf{Keystone} \\
\midrule
Privilege Model & Ring 3 enclaves & Secure Monitor & M-mode TCB \\
Openness & Closed & Closed & Open \\
Use Cases & Cloud security & Mobile security & Research, IoT \\
Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V \\
\bottomrule
\end{tabular}
\end{table}

@@

Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively. At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement. The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{Lee2019}.

@@

In RISC-V, the PMP mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode). Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode. The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode. This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,Lee2019}.
The basic hardware architecture of PMP within the RISC-V privilege hierarchy is illustrated in Figure~\ref{fig:pmp-arch}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=2.5cm,>=Stealth]
\tikzstyle{block} = [draw, minimum width=3cm, minimum height=2cm, align=center, font=\small]

@@

A typical RISC-V core supports a fixed number of PMP entries (commonly 8 or 16), each represented by two Control and Status Registers (CSRs): \texttt{pmpcfg} and \texttt{pmpaddr}. The \texttt{pmpcfg} register specifies the access permissions and address matching mode for each region, while \texttt{pmpaddr} defines the address bounds. RISC-V supports several region encoding modes:

\begin{itemize}
    \item \textbf{TOR (Top-of-Range):} Defines a region from the base address of the previous PMP entry up to the current address.
    \item \textbf{NAPOT (Naturally Aligned Power-Of-Two):} Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions.
    \item \textbf{NA4 (Naturally Aligned 4-byte):} Encodes a minimal 4-byte memory region.
\end{itemize}

@@

Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation. For example, Cerberus~\cite{lee2022cerberus} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components. Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.
Table~\ref{tab:pmp-limitations} summarizes key limitations in PMP and highlights recent research directions aimed at overcoming them.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
\caption{Kyber-768 Key Exchange Algorithm.}
\label{fig:kyber_768}
\end{figure}

@@

\begin{table}[h]
\centering
\caption{Key and Ciphertext Sizes (bytes)}
\label{tab:kyber_sizes}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
\hline
Kyber-512 & 1632 & 800 & 768 \\
Kyber-768 & 2400 & 1184 & 1088 \\
Kyber-1024 & 3168 & 1568 & 1568 \\
RSA-3072 & 384 & 384 & 384 \\
RSA-15360 & 1920 & 1920 & 1920 \\
\hline
\end{tabular}
\end{table}

@@

Empirical comparisons further highlight Keystone’s competitive performance. TS‑Perf, a systematic benchmarking framework, measured execution time across SGX, TrustZone, and RISC‑V Keystone using identical binaries and workloads (e.g., internal APIs, matrix operations, memory and storage access). It reports how architectural differences manifest in performance across ecosystems \cite{turn0search5}. As shown in Figure~\ref{fig:bench-compare}, Keystone achieves near-native performance on CPU-bound workloads and maintains a balanced profile across memory and I/O operations when compared to SGX and TrustZone.

@@

To address these performance issues, researchers have proposed alternative memory models that relax or restructure the traditional enclave-host memory boundaries. One such innovation is Elasticlave, a shared-memory model designed for enclaves that reduces isolation costs while preserving strong security properties. Elasticlave achieves an order of magnitude better performance compared to traditional spatial isolation models, with overheads as low as 10\% versus 100× to 1000× for spatially isolated configurations \cite{yu2022elasticlave}. Similarly, Eleos introduces self-paging within enclaves, allowing the enclave itself to manage virtual memory. This eliminates host OS involvement in page fault handling and mitigates context-switching and cache pollution overheads, resulting in significantly improved runtime efficiency \cite{orenbach2023eleos}. A related approach, Cerberus, applies formal methods to build a provably secure enclave memory-sharing model on Keystone. It enables safe data sharing with external components without compromising security, all while reducing redundancy and memory copying overheads \cite{lee2022cerberus}.

Input/output (I/O) operations also suffer substantial performance penalties in TEE contexts due to the need for secure copying and context switching. In Keystone, file I/O performance as measured using IOZone demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput compared to execution in the Rich Execution Environment (REE). Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\% \cite{Lee2019}. These results are visualized in Figure~\ref{fig:io-throughput}, which compares enclave and REE performance across several I/O workloads. Similar trends have been reported for other platforms; TS-perf reveals that both OP-TEE and SGX-based file operations incur high latency due to reliance on ECALLs and limited in-enclave storage APIs \cite{suzaki2021tsperf}.

@@

An additional source of performance variability lies in runtime behavior and process scheduling. Observations from TS-perf experiments showed unexpected enclave behavior such as thread migration between CPU cores and incorrect CPU utilization reporting, particularly in OP-TEE and Keystone environments. These anomalies were traced to a lack of TEE-awareness in OS-level schedulers and, in Keystone's case, bugs in the Eyrie runtime's interrupt handling mechanism \cite{suzaki2021tsperf}.

Monitoring and profiling enclave performance in real-time remains a difficult task. TEEMon provides a lightweight performance monitoring framework designed specifically for TEE environments. Though it introduces between 5–17\% overhead, it enables continuous visibility into runtime behavior and facilitates the identification of performance bottlenecks that would otherwise be hidden \cite{krahn2020teemon}.

>>>>

RISC-V specifies three hierarchical privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{Lee2019}. Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the Trusted Computing Base (TCB)—within M-mode. The Security Monitor manages enclaves and enforces memory isolation, while the runtime handles routine memory management and access mediation. By anchoring the TCB at this highest privilege level, Keystone minimizes the trusted surface exposed to potential attackers.

A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10 \cite{keystone2025how}. PMP empowers M-mode software to define fine-grained access permissions for physical memory regions, specifying which lower privilege levels (U and S) may read, write, or execute code therein~\cite{keystone2025how}. This hardware-enforced mechanism forms the cornerstone of Keystone’s memory isolation strategy. Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{Lee2019}.

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.48\linewidth}
  \centering
  \begin{tikzpicture}[font=\small]
    \draw[fill=green!10] (0,2) rectangle (3,5) node[midway]{\textbf{Secure World}};
    \draw[fill=red!10] (5,2) rectangle (8,5) node[midway]{\textbf{Normal World}};
    \draw[fill=gray!20] (2,0) rectangle (6,1.5) node[midway]{\textbf{Secure Monitor}};
    \draw[->,thick] (3,3.5) -- (5,3.5) node[midway,above]{\textbf{SMC}};
  \end{tikzpicture}
  \caption{TrustZone's dual-world architecture with secure monitor calls (SMC)-managed transitions.}
  \label{fig:trustzone}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\linewidth}
  \centering
  \begin{tikzpicture}[font=\small]
    \node[above] at (3,4) {\textbf{Enclave Page Cache (EPC)}};
    \draw[thick] (0,0) rectangle (6,4);

@@

\begin{table}[htbp]
\centering
\caption{Comparison of three TEE implementations.}
\label{tab:tee-comparison}
\small
%\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{3cm}p{2.5cm}p{2.5cm}p{2.5cm}@{}}
\toprule
\textbf{Feature} & \textbf{Intel SGX} & \textbf{ARM TrustZone} & \textbf{Keystone} \\
\midrule
Privilege Model & Ring 3 enclaves & Secure Monitor & M-mode TCB \\
Openness & Closed & Closed & Open \\
Use Cases & Cloud security & Mobile security & Research, IoT \\
Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V + PMP \\
\bottomrule
\end{tabular}
\end{table}

@@

Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively. At the heart of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement. The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{Lee2019}.

@@

In RISC-V, the Physical Memory Protection (PMP) mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode). Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode. The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode. This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,Lee2019}. Conceptually, M-mode software acts as a trusted monitor or security supervisor, responsible for configuring isolation policies and mediating access between different software privilege levels.
The basic hardware architecture of PMP within the RISC-V privilege hierarchy is illustrated in Figure~\ref{fig:pmp-arch}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=3cm,>=Stealth]
\tikzstyle{block} = [draw, minimum width=3cm, minimum height=2cm, align=center, font=\small]

@@

A typical RISC-V core supports a fixed number of PMP entries (commonly 8 or 16) \cite{keystone2025how}, each represented by two Control and Status Registers (CSRs): \texttt{pmpcfg} and \texttt{pmpaddr}. The \texttt{pmpcfg} register specifies the access permissions and address matching mode for each region, while \texttt{pmpaddr} defines the address bounds. RISC-V supports several region encoding modes as shown in Table~\ref{tab:pmp-modes}, each with distinct trade-offs in terms of flexibility and granularity~\cite{riscvprivspec}.

% \begin{itemize}
%     \item \textbf{TOR (Top-of-Range):} Defines a region from the base address of the previous PMP entry up to the current address.
%     \item \textbf{NAPOT (Naturally Aligned Power-Of-Two):} Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions.
%     \item \textbf{NA4 (Naturally Aligned 4-byte):} Encodes a minimal 4-byte memory region.
% \end{itemize}

\begin{table}[htbp]
\centering
\caption{PMP address-matching modes.}
\label{tab:pmp-modes}
\begin{tabular}{@{}lp{9cm}@{}} % adjust 9cm to fit your layout
\toprule
\textbf{Mode} & \textbf{Description} \\ \midrule
TOR   (Top-of-Range) & Defines a region from the base address of the previous PMP entry up to the current address. \\
NAPOT (Naturally Aligned Power-of-Two) & Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions. \\
NA4   (Naturally Aligned 4-byte) & Encodes a minimal 4-byte memory region. \\
\bottomrule
\end{tabular}
\end{table}

@@

Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation. For example, Cerberus~\cite{lee2022cerberus} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components, while HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in firmware, demonstrating PMP's applicability beyond general-purpose computing. Beyond CPU-managed memory, \textbf{IOPMP} highlights that malicious or compromised I/O devices could also access physical memory, potentially undermining traditional PMP protections, making it essential to extend isolation mechanisms to cover I/O memory as well.
Table~\ref{tab:pmp-limitations} summarizes key limitations in PMP and highlights recent research directions aimed at overcoming them.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
\caption{Kyber-768 Key Exchange Algorithm \cite{Pathum_2025}}
\label{fig:kyber_768}
\end{figure}

@@

\begin{table}[h]
\centering
\caption{Key and Ciphertext Sizes (bytes)}
\label{tab:kyber_sizes}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccc}
\toprule
\textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
\midrule
\rowcolor{lightgray} Kyber-512 & 1632 & 800 & 768 \\
\rowcolor{lightgray} Kyber-768 & 2400 & 1184 & 1088 \\
\rowcolor{lightgray} Kyber-1024 & 3168 & 1568 & 1568 \\
\rowcolor{darkgray} RSA-3072 & 384 & 384 & 384 \\
\rowcolor{darkgray} RSA-15360 & 1920 & 1920 & 1920 \\
\bottomrule
\end{tabular}
\end{table}

@@

Empirical comparisons further highlight Keystone’s competitive performance. TS‑Perf \cite{suzaki2021tsperf}, a systematic benchmarking framework, measured execution time across SGX, TrustZone, and RISC‑V Keystone using identical binaries and workloads (e.g., internal APIs, matrix operations, memory and storage access). It reports how architectural differences manifest in performance across ecosystems \cite{turn0search5}. As shown in Figure~\ref{fig:bench-compare}, Keystone achieves near-native performance on CPU-bound workloads and maintains a balanced profile across memory and I/O operations when compared to SGX and TrustZone.

@@

To address these performance issues, researchers have proposed alternative memory models that relax or restructure the traditional enclave-host memory boundaries. One such innovation is Elasticlave \cite{yu2022elasticlave}, a shared-memory model designed for enclaves that reduces isolation costs while preserving strong security properties. Elasticlave achieves an order of magnitude better performance compared to traditional spatial isolation models, with overheads as low as 10\% versus 100× to 1000× for spatially isolated configurations. Similarly, Eleos \cite{orenbach2023eleos} introduces self-paging within enclaves, allowing the enclave itself to manage virtual memory. This eliminates host OS involvement in page fault handling and mitigates context-switching and cache pollution overheads, resulting in significantly improved runtime efficiency. A related approach, Cerberus \cite{lee2022cerberus}, applies formal methods to build a provably secure enclave memory-sharing model on Keystone. It enables safe data sharing with external components without compromising security, all while reducing redundancy and memory copying overheads.

Input/output (I/O) operations also suffer substantial performance penalties in TEE contexts due to the need for secure copying and context switching. In Keystone, file I/O performance as measured using IOZone \cite{Lee2019} demonstrates a 36.2\% decrease in write throughput and a 40.9\% reduction in read throughput compared to execution in the Rich Execution Environment (REE). Repetitive operations, such as record rewriting, exhibit even greater overhead—up to 55.1\%. These results are visualized in Figure~\ref{fig:io-throughput}, which compares enclave and REE performance across several I/O workloads. Similar trends have been reported for other platforms; TS-perf reveals that both Open Portable Trusted Execution Environment (OP-TEE) and SGX-based file operations incur high latency due to reliance on ECALLs and limited in-enclave storage APIs \cite{suzaki2021tsperf}.

@@

An additional source of performance variability lies in runtime behavior and process scheduling. Observations from TS-perf experiments showed unexpected enclave behavior such as thread migration between CPU cores and incorrect CPU utilization reporting, particularly in Open Portable Trusted Execution Environment (OP-TEE) and Keystone environments. These anomalies were traced to a lack of TEE-awareness in OS-level schedulers and, in Keystone's case, bugs in the Eyrie runtime's interrupt handling mechanism \cite{suzaki2021tsperf}.

Monitoring and profiling enclave performance in real-time remains a difficult task. TEEMon \cite{krahn2020teemon} provides a lightweight performance monitoring framework designed specifically for TEE environments. Though it introduces between 5–17\% overhead, it enables continuous visibility into runtime behavior and facilitates the identification of performance bottlenecks that would otherwise be hidden.
